---
title: "A First Look"
description: |
  An initial look at the raw data, with an eye towards preprocessing.
author:
  - name: Kris Sankaran
    affiliation: UW Madison
date: "`r Sys.Date()`"
#output: distill::distill_article
output: pdf_document
params:
  data_dir: "/home/jovyan/data/"
  data_links: "https://uwmadison.box.com/shared/static/khd0plylp0o2tyhxlpizjj9rz7n5czya.csv"
---

1. In this notebook, we take a first look at the glacier data. We'll download
the raw annotations and images, using visualization to gain familiarity with the
structure of the problem.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

2. We'll be using the R packages below. `dplyr`, `ggplot2`, and `tidyr` are part
of R's tidy data ecosystem. `gdalutils`, `RStoolbox`, `raster` and `tmap` are
tools for spatial data manipulation and visualization.

```{r}
library("RStoolbox")
library("dplyr")
library("gdalUtils")
library("ggplot2")
library("tidyr")
library("tmap")
library("raster")
source("data.R")
theme_set(theme_bw())
```

3. In the block below, we'll create a directory structure to store the data in
our analysis. If you're working on your own computer, you can change the
directory to which files are written by modifying the `params` block in the
header of this rmarkdown file.

```{r}
x_dir <- file.path(params$data_dir, "tiles")
y_dir <- file.path(params$data_dir, "vector_data")
dir.create(x_dir, recursive = TRUE)
dir.create(y_dir)
```
4. Next, we download the actual imagery. The data are currently stored on a UW
Madison Box folder. We first download a file containing all the raw data
links^[It's very meta.], and then loop over just those links. This step can take
some time, because each satellite image is about 3GB.

```{r}
# index file, giving links to all the imagery
f <- tempfile()
download.file(params$data_links, f)
data_links <- read.csv(f, stringsAsFactors = FALSE)

# download all linked data
for (i in seq_len(nrow(data_links))) {
  download.file(
    data_links$link[i], 
    file.path(x_dir, data_links$name[i]),
    method = "libcurl",
    quiet = TRUE
  )
}
```

5. Often, we'll want to find the image associated with a particular point on the
earth. In the current form, this is quite hard to do: we'd have to check image
by image, until we found the one that had the geographic coordinate that we're
looking for. A much more convenient format is something called a ``Virtual
Tiff.'' This indexes all the imagery we had originally downloaded, letting us
look up imagery just by specifying the geographic coordinates. This means we can
hide all the complexity of working with many imagery files and pretend we had a
single (very large) image.

```{r}
x_paths <- dir(x_dir, "*tiff", full.names = TRUE)
vrt_path <- file.path(x_dir, "region.vrt")
gdalbuildvrt(x_paths, vrt_path)
```
6. Time for our first visualization! The `glaciers.geojson` file demarcates the
boundaries for two types of glaciers that are common in the Hindu Kush Himalayas
region. The full dataset gives coordinates for glaciers across the whole region,
which spans many countries, making it a bit overwhelming to visualize. Instead,
let's pick one of 63 basins in the region and visualize its glacier boundaries.

```{r}
y <- read_sf(file.path(y_dir, "glaciers.geojson")) %>%
  filter(Sub_basin == "Dudh Koshi")
head(y)

tm_shape(y) +
  tm_fill(col = "Glaciers", palette = c("#93b9c3", "#4e326a"))
```

7. From a technical perspective, there are two points worth noting. First, we
were able to use dplyr's `filter` function on these glaciers, as if they were
stored in an ordinary data frame (they're in fact stored in a spatial data
frame). Being able to use concepts of tidy data manipulation in spatial problems
can make our lives much easier. Second, notice that we're using `tm_shape` from
the `tmap` package -- this package makes much nicer spatial data visualizations
than simply calling `plot()` from base R. It also implements a grammar of
graphics for spatial data, allowing us to layer on different visual encodings in
a flexible way.

8. Next, let's visualize the satellite imagery in that region. The satellite
images are not like ordinary images from a camera -- they have many sensors. In
our case, each pixel is associated with 15 measurements. For example, the block
below first plots the RGB colors associated with the region before showing a
composite that turns all the glaciers blue.

```{r}
band_names <- names(brick(vrt_path)) # vrt loses the original channel names
x <- read_subset(vrt_path, st_bbox(y), band_names)
ggRGB(x)
ggRGB(x, r = 5, g = 4, b = 2)
```

9. Notice the similarities with the outlines in the `glaciers.geojson` file. We
can more or less distinguish the clean ice and debris-covered glaciers from the rest of
the image. Looking a bit more carefully, though, we realize that there are lots
of areas that fall into the bluish regions but which *aren't* labeled as clean
ice glacier. We might hope that some of the other channels make the difference
more visible, but it seems like there might be some danger of false positives.
Also, it seems like the debris-covered glaciers are only a subtley different
color from the background -- their tendril-like shape is a much better visual
clue.

10. We've managed to use `tm_shape` and `ggRGB` to avoid having to convert the
raw data to standard R data frames. There are some situations, however, when it
can be useful to make that conversion. The code below provides an example that
extracts the slope channel from the satellite image and visualizes it using
standard ggplot2. It seems like the debris-covered glaciers are relatively flat.

```{r}
elevation <- as.data.frame(x, xy = TRUE) %>%
  dplyr::select(x, y, slope)

ggplot(elevation, aes(x = x, y = y)) +
  geom_raster(aes(fill = slope)) +
  scale_fill_gradient(low = "white", high = "black") +
  coord_fixed()
```
11. Let's make a few histograms, one for each satellite sensor. From this view,
it becomes clear that we're going to have to preprocess these data before
modeling: The ranges across sensors can vary substantially.

```{r}
x_df <- x[sample(nrow(x), 250), ] %>% # subset pixels
  as.data.frame()
x_longer <- x_df %>%
  pivot_longer(cols = everything())

ggplot(x_longer) +
  geom_histogram(aes(x = value)) +
  facet_wrap(~ name, scale = "free_x")
```

12. For our final view, let's make a scatterplot of a few channels against one
another. We'll bin using `geom_hex`, because otherwise the points overlap too
much. This visualization makes it clear how the two `B6` channels are nearly
copies of one another, so we can safely drop one of them from our visualization.

```{r}
ggplot(x_df, aes(x = B1, y = B2, fill = log(..count..))) +
  geom_hex(binwidth = 2) +
  scale_fill_viridis_c() +
  coord_fixed()

ggplot(x_df %>% filter(B1 != 255, B2 != 255), aes(x = B1, y = B2, fill = log(..count..))) +
  geom_hex(binwidth = 2) +
  scale_fill_viridis_c() +
  coord_fixed()

ggplot(x_df, aes(x = B6_VCID_1, y = B6_VCID_2, fill = log(..count..))) +
  geom_hex(binwidth = 2) +
  scale_fill_viridis_c() +
  coord_fixed()
```

13. We haven't made too many plots, but we've already gained some valuable
intuition about (i) the types of visual features that might be useful for
identifying the two types of glaciers and (ii) some properties of the data that,
if not properly accounted for, could be disastrous for any downstream modeling
(no matter how fancy the model).